The reference model approach in feature selection problems \vspace{.1in}

### 2019-06-05 paper/ref_approach.tex
language edits

Here we use simulated data in order to control the difficulty of the selection problem. We rely on the scheme \eqref{eq:simulated_data} using different levels of correlation ($\rho$) and number observation ($n$), precisely $p$ and $k$ are fixed respectively at $1000$ and $100$, whereas $n\in\{50,70,100\}$ and $\rho\in\{0.3,0.5\}$. Lower the correlation level and the number of observation are, more challenging the selection is. As mentioned, this data generation mechanism was already proposed by \cite{paper:projpred}, who also devise a reference model sufficiently good in predictions. It consists in a linear regression using the first five supervised principal components (SPCs) \citep{paper:original_spc, paper:piironen_spc} as follows:
\
\begin{equation}
\label{eq:ref_mod}
\begin{aligned}
    Y_{i}|&\boldsymbol{\beta},\sigma^{2},\boldsymbol{u}_{i} \overset{ind}{\sim} N(\boldsymbol{u}_{i}^{T}\boldsymbol{\beta},\sigma^{2}) \quad &i=1,..,n \\
    &\beta_{j}|\!\begin{aligned}[t] &\tau \overset{iid}{\sim} N(0,\tau^{2})\\
    &\tau \sim t_{4}^{+}(0,s_{max}^{-2}) 
    \end{aligned} &j=1,..,5 \\ 
    &\sigma \sim t_{3}^{+}(0,10) \\
\end{aligned}
\end{equation}
where $u_{ij}$ stays for the $j$-th SPC of observation $i$, and $s_{max}$ denotes the sample standard deviation of the largest SPC. All model are implemented in Stan \citep{paper:stan}. The normal means problem formulation is obtained as explained in Section \ref{comparison} by Fisher-transformation. We label as ``ref'' the reference approach, which consists in using the reference model on top of the computation of the z-values, and ``data'' the standard approach based on the observed data only. In Section \ref{shrinkage-signal} we analyse in the context of full-Bayes methods the different shrinkage and signal estimation provided by the use of a reference model or not. Section \ref{complete-selection} shows results for the complete selection procedures.

\hypertarget{shrinkage-signal}{%
\subsubsection{Shrinkage and signal analysis}\label{shrinkage-signal}}
In the framework of full-Bayes methods we analyse the effect of the reference model using two continuous hierarchical shrinkage priors: the Dirichlet-Laplace \citep{paper:dirichlet_laplace} and the regularised horseshoe prior \citep{paper:rhs}. Although the horseshoe+ prior by \cite{paper:horseshoe+} is a state-of-the-art prior for sparse signal analysis in the context of the normal means problem, we decided to include the regularised horseshoe due to its faster MCMC inference, since our purpose in this paper is not to study the best prior choice, but rather compare and show the benefits of using a reference model when carrying out variable selection. We used an uniform prior on $(1/p,1/2)$ for the Dirichlet hyperparameter of the Dirichlet-Laplace prior, whereas regarding the reguelarised horseshoe prior we followed the indication provided in \cite{paper:rhs}, setting the prior guess of the effective number of parameters equal to 1 to achieve the highest shrinkage. Yet we have not observed any sensibility of the results due to such a choice.

As already mentioned, continuous priors do not usually give an automatic selection procedure, in this Section we study the amount of shrinkage and the bias of the posterior median looking at the SESE and the SSE (see definitions \eqref{eq:SSE} and \eqref{eq:SESE}). In the case of the regularised horseshoe prior, we also take into account the posterior mean for the shrinkage factor for each parameter $\theta_{j}$, which for model \eqref{eq:normal_means_problem2} is defined as:
\
\begin{equation}
\kappa_{j}=\frac{1}{1+\tau^{2}\lambda_{j}^{2}}
\end{equation}
with $\tau$ and $\lambda_{j}$ respectively global and local shrinkage parameters, for further details see \cite{paper:rhs}.

@@

Figure \ref{fig:k} shows the posterior medians for the shrinkage factors of each parameter, averaged after 100 data simulations. Relevant and non-relevant features are respectively highlighted in red and blue. The ideal shrinkage would have $\kappa=0$ when the parameter is a signal, whereas $\kappa=1$ when it is noise. Respectively on the y-axis and on the x-axis are depicted the shrinkage factors using the reference model approach or not. If the reference model did not make any difference, we would expect the points to lie on the diagonal. Non-relevant features happen to lie on average on the diagonal for all the simulated scenario, meaning not evident benefits coming from the reference model. Although when considering the shrinkage amount for the relevant ones, we note that the points always lie under the diagonal, meaning that the reference model is able to shrink less the true signals. Such benefit is proportional to the difficulty of the problem, that is it increases with lower correlation and number of observations. Note that the perfect distinction of the two cluster of features (relevant and non-relevant) observable on both the marginal distributions of the shrinkage factors is due to the averaging over the 100 data simulations. In a single data realisation there would be more noise and the cluster would be closer and more overlapping.

@@

\hypertarget{complete-selection}{%
\subsubsection{Complete selection analysis}\label{complete-selection}}
As complete selection procedures we consider the control of the local false discovery rate \citep{paper:efron, book:efron}, the empirical Bayes median \citep{paper:EBmed} and the selection by posterior credible intervals. The control of the local false discovery rate is applied through the R-package \textit{locfdr} using splines with 7 degrees of freedom to fit the marginal density and selecting features with local false discovery rate under the value 0.2. Such a value is the common suggested one just for the fact that it corresponds to a Bayes factor large than 36, other values can be considered and in our experience do not affect the results of the comparison. The empirical Bayes median procedure is given by the R-package \textit{EbayesThresh} and consists in fitting a Bayesian model with a prior composed by a mixture of a delta in zero and a heavy-tailed distribution. We use, as suggested, a Laplace distribution resulting in a thresholding property, i.e. there exists a threshold value such that all the data under that threshold have posterior median equal to zero. Therefore the selection is done including those parameters whose posterior median is different from zero. The hyperparameter of the Laplace distribution and the mixing weight of the prior are estimated by marginal maximum likelihood. The selection by posterior credible intervals is done using the regularised horseshoe prior and selecting those features who do not include zero in the interval between the 0.05 and the 0.95 quantiles.
\\
Figure \ref{fig:sensitivity_vs_fdr} reports the average sensitivity (on the y-axis) versus the average false discovery rate (x-axis) after 100 data simulations for the different combination of $n$ and $\rho$. For each method (different colours) the square and the circle dots respectively correspond to using or not the reference approach. The best selection performance is on the top left corner, which means lowest false discovery rate and highest sensitivity. It is possible to note that regardless of the method used for the selection, the use of a reference model improves the quality diminishing the false discovery rate (left shifting) and/or augmenting the sensitivity (up shifting). In accordance with what we would expect, higher the number of observation and the correlation are, easier the selection is and thus the benefit of the reference model are less notable, since the unprocessed data already give enough information to identify the significant variables.

@@

Figure \ref{fig:stability} shows the estimates of the stability measure proposed by \cite{paper:stability} with 0.95 confidence intervals after 100 simulations. Such a measure takes in account the variability of the subset of the selected features at each simulation (originally at each bootstrap sample) modelling the selection of each variable as a Bernoulli process. Further details are available in their paper as asymptotic normal distribution, based on which confidence intervals are computed. The reference model helps improving the stability of the selection: again such benefit is notable when the problem is more difficult (small $n$ and $\rho$). We observe also less uncertainty in the stability estimates for the reference approach (i.e. the width of the 0.95 confidence intervals), which can be still connected to the overall stability of the procedure.

@@

We conclude our experiments using the real world dataset body fat, already described in the comparison between the projection predictive approach and the stepwise regression in Section \ref{projection}. Consistently with the previous comparisons, we design the variable selection by normal means problem. We add to the original data noisy uncorrelated covariates up to a total of 1000 features. These covariates are normally distributed and scaled as the original ones. We compute correlations between each variable and the target variable, that is the amount of fat, and transform them by Fisher transformation. The original assumption in order that \eqref{eq:fisher_transformation} holds is that the variables are jointly normally distributed. In our experience the normal approximation is still reasonable, but after rescaling by $\sqrt{n-3}$ we do not fix the variance to be one. The methods we compare are those used in Section \ref{complete-selection}, which are the control of the local false discovery rate (loc.fdr), the empirical Bayes median (EB.med) and the selection by posterior credible intervals at level 0.9 (ci.90). The maximum likelihood estimator is used to estimate the variance of the null hypothesis in the local false discovery rate \cite[see][Chap. 6]{book:efron}, whereas the median absolute deviation from zero for the empirical Bayes median method. The selection via posterior credible intervals is done using a regularised horseshoe prior on the parameters for the means, while a standard log-normal  prior for the variance. In order to vary the difficulty of the selection, we bootstrap subsamples of different sizes, going from $n=50$ up to $n=251$. Each time results are averaged over 100 bootstrap subsamples.

@@

Figure \ref{fig:bodyfat_sensitivity_vs_fdr} shows the sensitivity against the false discovery rate. Since we do not have a ground truth regarding the original covariates of the data, we believe it is reasonable to consider all of them relevant, at least at some degree. We thus consider as non-relevant all the artificially added ones. When $n=70$ we observe an increment in terms of sensitivity, also the false discovery rate slightly increase, but in a lower amount. In all other bootstrapped subsamples, the reference model improves both in terms of sensitivity and false discovery rate. We observe again that the benefits are more evident as the selection is more challenging (i.e. less number of observations).
\\
Figure \ref{fig:bodyfat_stability} shows the stability results, still using the measure provided by \cite{paper:stability}. The benefit of the reference model are here marginal, small improvements can be observed, but overall not very notable.

>>>>

Here we use simulated data in order to control the difficulty of the selection problem. We rely on the scheme \eqref{eq:simulated_data} using different levels of correlation ($\rho$) and number observation ($n$), precisely $p$ and $k$ are fixed respectively at $1000$ and $100$, whereas $n\in\{50,70,100\}$ and $\rho\in\{0.3,0.5\}$. Lower the correlation level and the number of observations are, more challenging the selection is. As mentioned, this data generation mechanism was already proposed by \cite{paper:projpred}, who also devise a reference model sufficiently good in predictions. It consists in a linear regression using the first five supervised principal components (SPCs) \citep{paper:original_spc, paper:piironen_spc} as follows:
\
\begin{equation}
\label{eq:ref_mod}
\begin{aligned}
    Y_{i}|&\boldsymbol{\beta},\sigma^{2},\boldsymbol{u}_{i} \overset{ind}{\sim} N(\boldsymbol{u}_{i}^{T}\boldsymbol{\beta},\sigma^{2}) \quad &i=1,..,n \\
    &\beta_{j}|\!\begin{aligned}[t] &\tau \overset{iid}{\sim} N(0,\tau^{2})\\
    &\tau \sim t_{4}^{+}(0,s_{max}^{-2}) 
    \end{aligned} &j=1,..,5 \\ 
    &\sigma \sim t_{3}^{+}(0,10) \\
\end{aligned}
\end{equation}
where $u_{ij}$ stays for the $j$-th SPC of observation $i$, and $s_{max}$ denotes the sample standard deviation of the largest SPC. All model are implemented in Stan \citep{paper:stan}. The normal means problem formulation is obtained as explained in Section \ref{comparison} by Fisher-transformation. We label as ``ref'' the reference approach, which consists in using the reference model on top of the computation of the z-values, and ``data'' the standard approach based on the observed data only. In Section \ref{shrinkage-signal} we analyse in the context of full-Bayes methods the different shrinkage and signal estimation provided by the use of a reference model or not. Section \ref{complete-selection} shows results for the complete selection procedures.

\hypertarget{shrinkage-signal}{%
\subsubsection{Shrinkage and signal analysis}\label{shrinkage-signal}}
In the framework of full-Bayes methods we analyse the effect of the reference model using two continuous hierarchical shrinkage priors: the Dirichlet-Laplace \citep{paper:dirichlet_laplace} and the regularised horseshoe prior \citep{paper:rhs}. Although the horseshoe+ prior by \cite{paper:horseshoe+} is a state-of-the-art prior for sparse signal analysis in the context of the normal means problem, we decided to include the regularised horseshoe due to its faster MCMC inference, since our purpose in this paper is not to study the best prior choice, but rather compare and show the benefits of using a reference model when carrying out variable selection. We used an uniform prior on the interval $(1/p,1/2)$ for the Dirichlet hyperparameter of the Dirichlet-Laplace prior, whereas regarding the reguelarised horseshoe prior we followed the indication provided in \cite{paper:rhs}, setting the prior guess of the effective number of parameters equal to 1 to achieve the highest shrinkage. Yet we have not observed any sensibility of the results due to such a choice.

Since, as already mentioned, continuous priors do not usually give an automatic selection procedure, in this Section we study the amount of shrinkage and the bias of the posterior median looking at the SESE and the SSE (see definitions \eqref{eq:SSE} and \eqref{eq:SESE}). In the case of the regularised horseshoe prior, we also take into account the posterior mean for the shrinkage factor for each parameter $\theta_{j}$, which for model \eqref{eq:normal_means_problem2} is defined as:
\
\begin{equation}
\kappa_{j}=\frac{1}{1+\tau^{2}\lambda_{j}^{2}}
\end{equation}
with $\tau$ and $\lambda_{j}$ respectively global and local shrinkage parameters, for further details see \cite{paper:rhs}.

@@

Figure \ref{fig:k} shows the posterior medians for the shrinkage factors of each parameter, averaged after 100 data simulations. Relevant and non-relevant features are respectively highlighted in red and blue. The ideal shrinkage would have $\kappa=0$ when the parameter is a signal, whereas $\kappa=1$ when it is noise. Respectively on the y-axis and on the x-axis are depicted the shrinkage factors using the reference model approach or not. If the reference model did not make any difference, we would expect the points to lie on the diagonal. Non-relevant features happen to lie on average on the diagonal for all the simulated scenario, meaning not evident benefits coming from the reference model. Although when considering the shrinkage amount for the relevant ones, we note that the points always lie under the diagonal, meaning that the reference model is able to shrink less the true signals. Such benefit is proportional to the difficulty of the problem, that is it increases with lower correlation and number of observations. Note that the perfect distinction of the two clusters of features (relevant and non-relevant) observable on both the marginal distributions of the shrinkage factors is due to the averaging over the 100 data simulations. In a single data realisation there would be more noise and the clusters would be closer and more overlapping.

@@

\hypertarget{complete-selection}{%
\subsubsection{Complete selection analysis}\label{complete-selection}}
As complete selection procedures we consider the control of the local false discovery rate \citep{paper:efron, book:efron}, the empirical Bayes median \citep{paper:EBmed} and the selection by posterior credible intervals. The control of the local false discovery rate is applied through the R-package \textit{locfdr} using splines with 7 degrees of freedom to fit the marginal density and selecting features with local false discovery rate under the value 0.2. Such a value is commonly suggested only because it corresponds to a Bayes factor large than 36, other values can be considered and in our experience do not affect the results of the comparison. The empirical Bayes median procedure is given by the R-package \textit{EbayesThresh} and consists in fitting a Bayesian model with a prior composed by a mixture of a delta in zero and a heavy-tailed distribution. We use, as suggested, a Laplace distribution resulting in a thresholding property, i.e. there exists a threshold value such that all the data under that threshold have posterior median equal to zero. Therefore the selection is done including those parameters whose posterior median is different from zero. The hyperparameter of the Laplace distribution and the mixing weight of the prior are estimated by marginal maximum likelihood. The selection by posterior credible intervals is done using the regularised horseshoe prior and selecting those features who do not include zero in the interval between the 0.05 and the 0.95 quantiles.
\\
Figure \ref{fig:sensitivity_vs_fdr} reports the average sensitivity (on the y-axis) versus the average false discovery rate (x-axis) after 100 data simulations for the different combination of $n$ and $\rho$. For each method (different colours) the square and the circle dots respectively correspond to using or not the reference approach. The best selection performance is on the top-left corner of each plot, meaning lowest false discovery rate and highest sensitivity. It is possible to note that regardless of the method used, the use of a reference model improves the goodness of the selection diminishing the false discovery rate (left shifting) and/or augmenting the sensitivity (up shifting). In accordance with what expected, higher the number of observations and the correlation are, easier the selection is and, thus, the benefits of the reference model are less notable, since the unprocessed data already give enough information to identify the significant variables.

@@

Figure \ref{fig:stability} shows the estimates of the stability measure proposed by \cite{paper:stability} with 0.95 confidence intervals after 100 simulations. Such a measure takes in account the variability of the subset of the selected features at each simulation (originally at each bootstrap sample), modelling the selection of each variable as a Bernoulli process. Further details are available in their paper as asymptotic normal distribution, based on which confidence intervals are computed. The reference model helps improving the stability of the selection: again, such benefit is notable when the problem is more difficult (small $n$ and $\rho$). We observe also less uncertainty in the stability estimates for the reference approach (i.e. the width of the 0.95 confidence intervals), which can be still connected to the overall stability of the procedure.

@@

We conclude our experiments using the real world dataset body fat, already described in the comparison between the projection predictive approach and the stepwise regression in Section \ref{projection}. Consistently with the previous comparisons, we design the variable selection by normal means problem. We add to the original data noisy uncorrelated covariates up to a total of 1000 features. These covariates are normally distributed and scaled as the original ones. We compute correlations between each variable and the target variable, that is the amount of fat, and transform them by Fisher transformation. The original assumption, in order that \eqref{eq:fisher_transformation} holds, is that the variables are jointly normally distributed. In our experience the normal approximation is still reasonable, but after rescaling by $\sqrt{n-3}$ we do not fix the variance to be one, but we estimate it from the data. The methods we compare are those used in Section \ref{complete-selection}: the control of the local false discovery rate (loc.fdr), the empirical Bayes median (EB.med) and the selection by posterior credible intervals at level 0.9 (ci.90). The maximum likelihood estimator is used to estimate the variance of the null hypothesis in the local false discovery rate \cite[see][Chap. 6]{book:efron}, whereas the median absolute deviation from zero for the empirical Bayes median method. The selection via posterior credible intervals is done using a regularised horseshoe prior on the parameters for the means, while a standard log-normal  prior for the variance. In order to vary the difficulty of the selection, we bootstrap subsamples of different sizes, going from $n=50$ up to $n=251$. Each time results are averaged over 100 bootstrap subsamples.

@@

Figure \ref{fig:bodyfat_sensitivity_vs_fdr} shows the sensitivity against the false discovery rate. Since we do not have a ground truth regarding the original covariates of the data, we believe it is reasonable to consider all of them relevant, at least at some degree. We thus consider as non-relevant all the artificially added ones. In almost all the bootstrapped subsamples, the reference model improves both in terms of sensitivity and false discovery rate. When $n=50$, we observe a worsening in terms of false discovery rate, yet in a lower amount compared to the gain in sensitivity.  We observe again that the benefits are more evident as the selection is more challenging (i.e. less number of observations).
\\
Figure \ref{fig:bodyfat_stability} shows the stability results, still using the measure provided by \cite{paper:stability}. The benefits of the reference model are here marginal: small improvements can be observed, but overall not very notable.
